Effects of Antenna Beam Chromaticity on Redshifted 21~cm Power Spectrum and Implications for Hydrogen Epoch of Reionization Array

### 2016-03-15 ForegroundSimulations/ms.tex
Trim text in Abstract and Introduction
From cbfc31c (fgsim):
====================

In ms.tex,

(*) Trim text in Abstract and Introduction

Unaccounted systematics from foregrounds and instruments prevent current experiments from detecting redshifted 21~cm signal from the Epoch of Reionization (EoR). In-depth understanding of these effects is critical for such experiments to succeed. We identify spectral structures (chromaticity) in the antenna power pattern as one such systematic that significantly affects all EoR measurements. Increase in chromaticity of the antenna beam causes the already bright foreground power to leak well beyond ideal limits and contaminate the otherwise clean EoR signal modes. A primary source of this chromaticity is reflections in the antenna-feed assembly and between structures in neighboring antennas. Using precise foreground simulations taking wide-field effects into account, we provide a framework to set cosmologically-motivated design specifications on these reflections to prevent further EoR signal degradation. We use the Hydrogen Epoch of Reionization Array (HERA) as an example. Using a simple and yet high dynamic range spectral weighting scheme, we demonstrate that EoR detection using HERA will not be impeded by such spectral systematics even in a conservative scenario that does not perform removal of foregrounds. We also show that all baselines in a 19-element HERA layout are capable of detecting EoR over a substantial observing window on the sky.

@@

The period in the history of the Universe characterized by the transition of neutral hydrogen in the intergalactic medium (IGM) to a fully ionized state due to the formation of radiating objects such as the first stars and galaxies is referred to as the Epoch of Reionization (EoR). This is an important period of nonlinear growth of matter density perturbations and astrophysical evolution leading to the large scale structure observed currently in the Universe. And yet, this period in the Universe's history has remained poorly probed to date.

The redshifted neutral hydrogen from the IGM in this epoch has been identified to be one of the most promising and direct probes of the EoR \citep{sun72,sco90,mad97,toz00,ili02}. Numerous experiments using low frequency radio telescopes targeting the redshifted 21~cm line from the spin-flip transition of H{\sc i} have become operational such as the Murchison Widefield Array \citep[MWA;][]{lon09,bow13,tin13}, the Donald~C.~Backer Precision Array for Probing the Epoch of Reionization \citep[PAPER;][]{par10}, the Low Frequency Array \citep[LOFAR;][]{van13} and the Giant Metrewave Radio Telescope EoR experiment \citep[GMRT;][]{pac13}. Many of these instruments have sufficient sensitivity for a statistical detection of the EoR signal by estimating the spatial power spectrum of the redshifted H{\sc i} spin temperature fluctuations \citep{bea13,thy13}. These instruments are intended to be precursors and pathfinders to the next generation of low frequency radio observatories such as the Hydrogen Epoch of Reionization Array\footnote{\url{http://reionization.org/}} \citep[HERA;][]{deb16} and the Square Kilometre Array\footnote{\url{https://www.skatelescope.org/}} (SKA). These next-generation instruments will advance the capability from a mere statistical detection of the signal to a direct three-dimensional tomographic imaging of H{\sc i} from the EoR.

The most significant challenge to low frequency EoR observations arises from the extremely bright Galactic and extragalactic foreground synchrotron emission which are $\sim 10^4$ times stronger than the desired EoR signal \citep{dim02,ali08,ber09,ber10,gho12}. All the current and future instruments rely on the inherent differences in spatial isotropy and spectral smoothness between the EoR signal and the foregrounds to extract the EoR power spectrum \citep[see, e.g.,][]{fur04b,mor04,zal04,san05,fur06,mcq06,mor06,wan06,gle08}.

When expressed in the coordinate system of power spectrum measurements described by the three-dimensional wavenumber ($k$), the foreground emission is restricted to a wedge-shaped region commonly referred to as the {\it foreground wedge} \citep{bow09,liu09,liu14a,liu14b,dat10,liu11,gho12,mor12,par12b,tro12,ved12,dil13,pob13,thy13,dil14} bounded by the horizon limits \citep{par12b}, whereas the EoR signal has spherical symmetry due to its isotropy which appears elongated along line-of-sight $k$-modes due to peculiar velocity effects when dominated by matter density perturbations during early stages of reionization. The region of $k$-space excluding the {\it foreground wedge} is commonly referred to as the {\it EoR window} since it is expected to be relatively free of foreground contamination.

However, despite attempts to exploit this knowledge about differences in spectral and spatial symmetry in lengthy observations, analysis from the first-generation experiments has at best only led to weak constraints on reionization models \citep{ali15,dil15,pob15}. This is because their sensitivity is primarily limited by instrumental and foreground systematics. The extreme dynamic range required to suppress foreground and instrument systematics demands high precision modeling of foregrounds as observed by modern wide-field instruments. One such study was undertaken by \citet{thy15a,thy15b}. Their studies of effects of wide-field measurements of EoR power spectra have demonstrated the {\it pitchfork} effect wherein foreshortening of baselines causes a prominent enhancement of foreground power near the horizon limits of the {\it foreground wedge} as well as in the contamination beyond. It highlighted the importance of controlling the response of wide-field instruments near the horizon. The need for precise subtraction of foregrounds in wide-field measurements was also highlighted in \citet{pob16}.

In this paper, we explore yet another phenomenon that inherently extends foreground power beyond the {\it wedge}, namely, that arising from the chromaticity of the antenna power pattern. While highlighting the importance of such instrumental systematics on future EoR experiments, we demonstrate that the current HERA instrument design will exceed these specifications to detect the EoR signal with high significance. This will hold when data is limited by foregrounds and systematics even in the most conservative scenario which involves neither foreground subtraction nor optimal analysis techniques but only relies on a simple spectral weighting scheme.

This paper is organized as follows. \S\ref{sec:HERA} introduces the HERA instrument. A brief summary of the delay spectrum technique used extensively in this analysis and the recently confirmed {\it pitchfork} effect are presented in \S\ref{sec:delay-spectrum}. \S\ref{sec:sim} describes foreground simulations including antenna beam pattern, all-sky foreground models and two independent EoR models. \S\ref{sec:beam-chromaticity} investigates the effects of chromaticity of antenna beam on the resulting delay power spectrum and the cosmologically motivated constraints it places on reflections in the instrument. EoR-foreground dynamic range of HERA under simple {\it foreground avoidance} criteria are demonstrated in \S\ref{sec:eor-sensitivity}. Our findings are summarized in \S\ref{sec:summary}.

>>>>

Unaccounted systematics from foregrounds and instruments prevent current experiments from detecting redshifted 21~cm signal from the Epoch of Reionization (EoR). We identify spectral structures (chromaticity) in the antenna power pattern as one such systematic that significantly affects all EoR measurements. Increase in chromaticity of the antenna beam causes the already bright foreground power to leak well beyond ideal limits and contaminate the otherwise clean EoR signal modes. A primary source of this chromaticity is reflections in the antenna-feed assembly and between structures in neighboring antennas. Using precise foreground simulations taking wide-field effects into account, we provide a framework to set cosmologically-motivated design specifications on these reflections to prevent further EoR signal degradation. We use the Hydrogen Epoch of Reionization Array (HERA) as an example. Using a simple and yet high dynamic range spectral weighting scheme, we demonstrate that EoR detection using HERA will not be impeded by such spectral systematics even in a conservative scenario that does not perform removal of foregrounds. We also show that all baselines in a 19-element HERA layout are capable of detecting EoR over a substantial observing window on the sky.

@@

% The period in the history of the Universe characterized by the transition of neutral hydrogen in the intergalactic medium (IGM) to a fully ionized state due to the formation of radiating objects such as the first stars and galaxies is referred to as the Epoch of Reionization (EoR). This is an important period of nonlinear growth of matter density perturbations and astrophysical evolution leading to the large scale structure observed currently in the Universe. And yet, this period in the Universe's history has remained poorly probed to date.

The Epoch of Reionization (EoR) is an important period of nonlinear growth of matter density perturbations and astrophysical evolution leading to the large scale structure observed currently in the Universe. The redshifted neutral hydrogen from the IGM in this epoch has been identified to be one of the most promising and direct probes of the EoR \citep{sun72,sco90,mad97,toz00,ili02}.

Numerous experiments using low frequency radio telescopes targeting the redshifted 21~cm line from the spin-flip transition of H{\sc i} in this epoch have become operational such as the Murchison Widefield Array \citep[MWA;][]{lon09,bow13,tin13}, the Donald~C.~Backer Precision Array for Probing the Epoch of Reionization \citep[PAPER;][]{par10}, the Low Frequency Array \citep[LOFAR;][]{van13} and the Giant Metrewave Radio Telescope EoR experiment \citep[GMRT;][]{pac13}. Many of these instruments have sufficient sensitivity for a statistical detection of the EoR signal by estimating the spatial power spectrum of the redshifted H{\sc i} spin temperature fluctuations \citep{bea13,thy13}. These instruments are intended to be precursors and pathfinders to the next generation of low frequency radio observatories such as the Hydrogen Epoch of Reionization Array\footnote{\url{http://reionization.org/}} \citep[HERA;][]{deb16} and the Square Kilometre Array\footnote{\url{https://www.skatelescope.org/}} (SKA). These next-generation instruments will advance the capability from a mere statistical detection of the signal to a direct three-dimensional tomographic imaging of H{\sc i} from the EoR.

The most significant challenge to low frequency EoR observations arises from the extremely bright Galactic and extragalactic foreground synchrotron emission which are $\sim 10^4$ times stronger than the desired EoR signal \citep{dim02,ali08,ber09,ber10,gho12}. However, there are inherent differences in spatial isotropy and spectral smoothness between the EoR signal and the foregrounds \citep[see, e.g.,][]{fur04b,mor04,zal04,san05,fur06,mcq06,mor06,wan06,gle08}. % All the current and future instruments rely on the inherent differences in spatial isotropy and spectral smoothness between the EoR signal and the foregrounds to extract the EoR power spectrum \citep[see, e.g.,][]{fur04b,mor04,zal04,san05,fur06,mcq06,mor06,wan06,gle08}.

When expressed in the coordinate system of power spectrum measurements described by the three-dimensional wavenumber ($k$), the foreground emission is restricted to a wedge-shaped region commonly referred to as the {\it foreground wedge} \citep{bow09,liu09,liu14a,liu14b,dat10,liu11,gho12,mor12,par12b,tro12,ved12,dil13,pob13,thy13,dil14} bounded by the horizon limits \citep{par12b}, whereas the EoR signal has spherical symmetry due to its isotropy. % which appears elongated along line-of-sight $k$-modes due to peculiar velocity effects when dominated by matter density perturbations during early stages of reionization. 
The region of $k$-space excluding the {\it foreground wedge} is commonly referred to as the {\it EoR window} since it is expected to be relatively free of foreground contamination.

Despite attempts to exploit this knowledge about differences in spectral and spatial symmetry, first-generation experiments have at best only led to weak constraints on reionization models \citep{ali15,dil15,pob15}. This is because their sensitivity is primarily limited by instrumental and foreground systematics. The extreme dynamic range required to suppress foreground and instrument systematics demands high precision modeling of foregrounds as observed by modern wide-field instruments.

% One such study was undertaken by \citet{thy15a,thy15b}. Their studies of effects of wide-field measurements of EoR power spectra have demonstrated the {\it pitchfork} effect wherein foreshortening of baselines causes a prominent enhancement of foreground power near the horizon limits of the {\it foreground wedge} as well as in the contamination beyond. It highlighted the importance of controlling the response of wide-field instruments near the horizon. The need for precise subtraction of foregrounds in wide-field measurements was also highlighted in \citet{pob16}.

In this paper, we explore a phenomenon that inherently extends foreground power beyond the {\it wedge}, namely, that arising from the chromaticity of the antenna power pattern. While highlighting the importance of such instrumental systematics on future EoR experiments, we demonstrate that the current HERA instrument design will exceed these specifications to detect the EoR signal with high significance. This will hold true when data is limited by foregrounds and systematics even in the most conservative scenario which involves neither foreground subtraction nor optimal analysis techniques but only relies on a simple spectral weighting scheme.

This paper is organized as follows. \S\ref{sec:HERA} describes the HERA instrument. A brief summary of the delay spectrum technique used extensively in this analysis and the recently confirmed wide-field instrument effects are presented in \S\ref{sec:delay-spectrum}. \S\ref{sec:sim} describes foreground simulations including antenna beam pattern, all-sky foreground models and two independent EoR models. \S\ref{sec:beam-chromaticity} investigates the effects of chromaticity of antenna beam on the resulting delay power spectrum and the cosmologically motivated constraints it places on reflections in the instrument. EoR-foreground dynamic range of HERA under simple {\it foreground avoidance} criteria are demonstrated in \S\ref{sec:eor-sensitivity}. Our findings are summarized in \S\ref{sec:summary}.
